---
title: "R: Lectura de datos"
author: "Abraham Jaimes"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Datos

## Tamaño de los datos

```{r}
nrow(iris)
ncol(iris)
dim(iris)
nrow(ggplot2::diamonds)
```

## Guardado de datos 

```{r}
write.csv(iris, "~/Descargas/iris.csv", row.names = FALSE)
write.csv(ggplot2::diamonds, "~/Descargas/diamonds.csv", row.names = FALSE)
zip("~/Descargas/iris.zip", normalizePath("~/Descargas/iris.csv"))
```

# Lectura

## Base R

```{r}
read.csv("~/Descargas/iris.csv")
```

## `data.table`

```{r}
library(data.table)
```

```{r}
fread("~/Descargas/iris.csv")
```

## Benchmark

```{r}
library(microbenchmark)
```

### Pocos datos

```{r}
microbenchmark(base = read.csv("~/Descargas/iris.csv"),
               data_table = fread("~/Descargas/iris.csv"), 
               times = 20)
```

### Muchos datos

```{r}
microbenchmark(base = read.csv("~/Descargas/diamonds.csv"),
               data_table = fread("~/Descargas/diamonds.csv"),
               times = 20)
```

### Ejemplos avanzados `fread`

#### input

```{r}
fread(input = "date, count
2022-12-27 00:00:00,     0
2022-12-27 00:01:00,     0
2023-01-03 00:00:00,     0
2023-01-03 00:01:00,     0
2023-01-10 00:00:00,     0
2023-01-10 00:01:00,     0
2023-01-17 00:00:00,     1
2023-01-17 00:01:00,     0")
```

```{r}
fread(text = "date, count
2022-12-27 00:00:00,     0
2022-12-27 00:01:00,     0
2023-01-03 00:00:00,     0
2023-01-03 00:01:00,     0
2023-01-10 00:00:00,     0
2023-01-10 00:01:00,     0
2023-01-17 00:00:00,     1
2023-01-17 00:01:00,     0")
```

```{r}
fread("unzip -p ~/Descargas/iris.zip")
```

```{r}
fread(cmd = "unzip -p ~/Descargas/iris.zip")
```

```{r}
fread("~/Descargas/iris.csv")
```

```{r}
fread(file = "~/Descargas/iris.csv")
```

#### cmd

```{r}
fread(cmd = paste("grep", "setosa", "~/Descargas/iris.csv"))
```

```{r}
fread(cmd = "unzip -p ~/Descargas/iris.zip")
```

#### `nrows` y `select`

```{r}
fread("~/Descargas/iris.csv", nrows = 3, select = c(1, 2))
fread("~/Descargas/iris.csv", nrows = 3, select = c("Sepal.Length", "Species"))
```

#### `skip`

```{r}
fread("~/Descargas/iris.csv", skip = 100)
fread("~/Descargas/iris.csv", skip = "Sepal.Length")
```

## Tidyverse

### readxl

```{r}
library(readxl)
```

```{r}
datasets <- readxl_example("datasets.xlsx")
read_excel(datasets)

# Specify sheet either by position or by name
read_excel(datasets, 2)
read_excel(datasets, "mtcars")

# Skip rows and use default column names
read_excel(datasets, skip = 148, col_names = FALSE)

# Recycle a single column type
read_excel(datasets, col_types = "text")

# Specify some col_types and guess others
read_excel(datasets, col_types = c("text", "guess", "numeric", "guess", "guess"))

# Accomodate a column with disparate types via col_type = "list"
df <- read_excel(readxl_example("clippy.xlsx"), col_types = c("text", "list"))
df
df$value
sapply(df$value, class)

# Limit the number of data rows read
read_excel(datasets, n_max = 3)

# Read from an Excel range using A1 or R1C1 notation
read_excel(datasets, range = "C1:E7")
read_excel(datasets, range = "R1C2:R2C5")

# Specify the sheet as part of the range
read_excel(datasets, range = "mtcars!B1:D5")

# Read only specific rows or columns
read_excel(datasets, range = cell_rows(102:151), col_names = FALSE)
read_excel(datasets, range = cell_cols("B:D"))

# Get a preview of column names
names(read_excel(readxl_example("datasets.xlsx"), n_max = 0))

# exploit full .name_repair flexibility from tibble

# "universal" names are unique and syntactic
read_excel(
  readxl_example("deaths.xlsx"),
  range = "arts!A5:F15",
  .name_repair = "universal"
)

# specify name repair as a built-in function
read_excel(readxl_example("clippy.xlsx"), .name_repair = toupper)

# specify name repair as a custom function
my_custom_name_repair <- function(nms) tolower(gsub("[.]", "_", nms))
read_excel(
  readxl_example("datasets.xlsx"),
  .name_repair = my_custom_name_repair
)

# specify name repair as an anonymous function
read_excel(
  readxl_example("datasets.xlsx"),
  sheet = "chickwts",
  .name_repair = ~ substr(.x, start = 1, stop = 3)
)
```

## dplyr

```{r}
library(dplyr)
```

```{r}
mtcars1 <- mtcars
```

```{r}
rownames(mtcars1)
```

```{r}
mtcars1$modelo <- rownames(mtcars1)
```

```{r}
mtcars2 <- mtcars
```

## mutate: Agregar columnas

```{r}
mtcars2 <- mutate(mtcars2, modelo = rownames(mtcars2), modelo_lower = tolower(modelo))
```

## pipe

```{r}
log10(exp(sin(0)))
0 %>% sin()
0 |> sin()
```

```{r}
mtcars2 <- mtcars2 %>% 
  mutate(modelo = rownames(mtcars2), modelo_lower = tolower(modelo))
```

## filter: Filtrar 

```{r}
mtcars2[mtcars2$mpg > 20, ]
mtcars2[mtcars2$mpg > 20 & mtcars2$hp > 100, ]
```


```{r}
filter(mtcars2, mpg > 20)
```

```{r}
mtcars2 %>% filter(mpg > 20, hp > 100)
```

```{r}
microbenchmark::microbenchmark(base = mtcars2[mtcars2$mpg > 20 & mtcars2$hp > 100, ],
                               dplyr = mtcars2 %>% filter(mpg > 20, hp > 100))
```

```{r}
mtcars2 %>% 
  mutate(modelo = rownames(mtcars2)) %>%
  filter(mpg > 20, hp > 100)
```

## select: Seleccionar columnas

```{r}
select(mtcars2, mpg, cyl)
```

```{r}
mtcars2[, c("mpg", "cyl")]
```

```{r}
mtcars2 %>% 
  mutate(modelo = rownames(mtcars2)) %>%
  filter(mpg > 20, hp > 100) %>%
  select(mpg, cyl, hp) %>%
  mutate(double_hp = 2 * hp) %>%
  #mutate(double_hp = 2 * hp, hp = NULL) %>%
  select(mpg, cyl, double_hp)
```

## arrange: Reordernar

```{r}
arrange(mtcars2, cyl, -mpg)
```

## summarise: Resumen de los datos

```{r}
mtcars2 %>% 
  summarise(mean_hp = mean(hp), max_hp = max(hp), min_hp = min(hp))
```

```{r}
mtcars2 %>%
  arrange(-hp)
```

## group_by: Agrupar

```{r}
mtcars %>%
  group_by(cyl) %>%
  summarise(mean_hp = mean(hp), median_hp = median(hp))
```


```{r}
library(ggplot2)
```

```{r}
mtcars2 %>%
  filter(hp >= 150) %>%
  mutate(am = as.logical(am)) %>%
  ggplot(aes(x = am, y = mpg)) + geom_boxplot() 
```

## data.table

```{r}
dt_mtcars <- as.data.table(mtcars)
```

### Filtrar

```{r}
filter(mtcars2, mpg > 20) # dplyr
dt_mtcars[mpg >= 20] # data.table
```

### Agregar columnas

```{r}
mutate2 <- mutate(mtcars2, double_hp = 2 * hp) # dplyr
dt_mtcars[, double_hp := 2 * hp] # data.table
```

### Summary

```{r}
summarise(mtcars2, mean_hp = mean(hp), max_hp = max(hp), min_hp = min(hp))
dt_mtcars[, .(mean_hp = mean(hp), max_hp = max(hp), min_hp = min(hp))]
```





```{r}
inicia_con <- function(x) substr(x, 1, 1)
iris %>%
  filter(inicia_con(Species) == "v")
```

```{r}
iris %>%
  filter(2 * Sepal.Length == 10.8)
```

```{r}
iris %>%
  select(contains("pe")) %>%
  head()
```


```{r}
dt_iris <- iris
cols <- grep("pe", colnames(dt_iris), ignore.case = TRUE)
dt_iris[, cols]
```



```{r}
iris <- iris %>%
  mutate(Species = as.character(Species))

str(iris)

iris %>%
  arrange(Species)

iris$Species[order(iris$Species, decreasing = TRUE)]

index <- order(iris$Species, decreasing = TRUE)
iris[index, ]

iris %>%
  slice(50)
```
